Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-06-21 demystify.tex
updates

International students, especially those from less well-known universities and smaller countries, often wonder if they should apply to a CS PhD program in the US. Specifically, they are worried about the \emph{difficulty} of getting admitted and the \emph{cost} of studying in the US. This section addresses these concerns.

First, applying to a good US university \emph{should not} be any harder than at schools in other countries. In fact, it might even be more flexible since CS PhD in the US \emph{do not} require having an MS or a topic/proposal/adviser in advance. If you believe you have a chance in other countries, e.g., South Korea, Singapore, Germany, UK, Japan, and Australia, then you will surely have a chance in the US as well. \autoref{sec:non-us-differences} compares CS PhD study in the US to other countries.

@@

\begin{commentbox}[Vu:]
  One of the reasons I created this document is that my colleagues at GMU are interested in recruiting Vietnamese students and are surprised when seeing very few applications in Vietnam (see \autoref{sec:ack}). Each year our CS PhD program receives 500+ PhD applications, most of which are international but only 5--6 are from Vietnam. In general the number of CS PhD applications from Vietnam to US universities is very low and more would be welcome.
\end{commentbox}

@@

The guideline can be useful for all students, but it is primarily intended for \textbf{international students} from \textbf{smaller countries and less well-known universities} (similarly, for domestic students from US universities with no PhD programs or limited research experiences). 
Students from top schools with strong research cultures and experiences might already know most of the information given here.
My goal is thus to level the playing field by providing information that is not readily available to less privileged students. 
I also hope to encourage more students with such backgrounds to apply to CS PhD programs in the US.

While this document may not fully meet the needs of students aiming for top-10 schools, which typically have more stringent requirements (e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU), it should be valuable for those targeting other reputable R1 institutions. It is certainly useful for students aiming for GMU!

@@

\section*{Additional Resources}
\begin{itemize}
  \item \href{https://parentheticallyspeaking.org/articles/us-cs-phd-faq/}{Getting a Computer Science PhD in the USA} by Shriram Krishnamurthi
  \item  \href{https://pg.ucsd.edu/PhD-application-tips.htm}{Blog post on PhD Admission} by Philip Guo
  \item   \href{https://talkingtorobots.com/yonatanbisk.html}{PhD Admission post} by Yonatan Bisk (customized for CMU admission)
\end{itemize}

@@

After you submit your PhD application (usually around December of year $X-1$ for the Fall admission of year $X$), it will first be screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters? Usually, this screening process is done through a central university system, i.e., not by CS faculty. If you pass this screen, your application will be forwarded to the CS department for further evaluation. If you don't,  the system will tell you what is missing and what you need to do. So pay attention to your email and check your application status regularly.

\begin{commentbox}[Hakan:]
  At GMU, for full consideration, students should make sure to submit \textbf{ALL} required documents by the application deadline, and should never assume that some required documents (such as official TOEFL scores or official diplomas/transcripts) will be waived by the admissions office. If something is listed and not marked as ``optional'', it is mandatory and they should plan for submitting all those.
\end{commentbox}

Then, applications are reviewed by a \emph{PhD admission committee} (\textbf{adcom}) that consists of faculty members in CS\footnote{In some cases the committee can involve affiliated faculty from different disciplines.}. These adcom members have a wide range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depend on the department size. At GMU, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review 25--30 applications. Note that most large schools, including GMU, have separate adcoms for MS programs.

@@

At GMU, we usually decide that a full-time PhD candidate is either (i) admitted with funding (\autoref{sec:funding}) or (ii) rejected. In other words, in most cases, we either
admit you with full funding or reject your application. In some rare cases, we admit
without funding because you have funding on your own, e.g.,
supported by your government or having external fellowships. We justify
our decision with a summary of your application, where we list
strengths, e.g., a well-known school, and weaknesses, e.g., weak
LORs.

@@

\begin{commentbox}[Craig:]
  GMU and many other universities allow you to upload your published papers and other writing samples. In many cases, even if the papers were not published at top places, we can still determine their quality by simply skimming over the paper.
\end{commentbox}

@@

\begin{commentbox}[Vu:]
  Sometimes PhD adcom in the US will share a document such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Foreign-Top-Schools}{this one}, which lists the top schools in several countries. We also ask other faculty and students if we think they know about the place.  For example, when I was a postdoc at UMD, members of their CS PhD adcom asked me to evaluate applicants from Vietnam.  During my time at UNL and now here at GMU, I have looked at Vietnamese applications (whether they are assigned to me or not) and provided input to their reviewers, e.g., X is the top tech school in Vietnam and so it should be \emph{top} instead of \emph{unknown} foreign, which makes a huge difference.
\end{commentbox}
\begin{commentbox}[Deepak:]
  If an applicant is anxious about their school not being known outside their country, they can provide information about their school and department, with independent sources where such information can be verified.
\end{commentbox}

@@

\paragraph{English Tests (IELTS, TOEFL)} Unless your degrees are from specific countries such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{these}, you will need to
take standardized English tests.

@@

\begin{commentbox}[Vu:]
  The minimum for GMU (being above this might not mean much, but below is a \red{red flag}).
  \begin{itemize}
    \item GPA: $\ge 3.0$ in your undergrad (but we also consider the rank/prestige of your school)
    \item GRE: not required
    % \item but if you want to use it, then we expect a total (V+Q) of $\ge 311$ (with a $\ge 157$ Q) and A $\ge 3.0-3.5$.
    \item English requirement tests (one of the below)
          \begin{itemize}
            \item TOEFL: 88 pts in total AND $\ge 20$ points in each subsection OR
            \item IELTS: $\ge 6.5$ OR
            \item DuoLingo Graduate English: $\ge 120$ OR
            \item Pearson Test of Academic English: $\ge 67$
          \end{itemize}
  \end{itemize}
\end{commentbox}

@@

\begin{commentbox}[Vu:]
  At GMU, faculty are encouraged to interview candidates. For very strong candidates, the interview is actually to recruit them.  In some cases a faculty interviews a candidate that they see potential and want to advocate for their admission. Without the interview, such applications may be more likely to be rejected.
  \tcblower
  In short, getting an interview is a good sign; it means that someone is considering you. If we are not interested in your application, we will not proceed with an interview.
\end{commentbox}

@@

\paragraph{Not getting interviews} While it is generally good to get an interview, not getting one \emph{does not} mean you're out.  Many programs do not have the tradition of interviewing applicants. For example, at GMU, most admitted students with TA (\autoref{sec:ta}) do not go through interviews.

@@

\begin{commentbox}[Vu:]
  GMU has \emph{Virtual} Open House (VOH), e.g., \url{https://cs-gmu.github.io/cs-phd-voh-s23/}. We invite all admitted PhD students to the VOH through Zoom to learn about the CS program, the department, GMU, and the DC area in general. Students also get opportunities to chat with professors and current students.
\end{commentbox}

@@

\paragraph{Low Stipend?} Students often complain their stipend is too low. However, in most cases it is not bad, and you can live comfortably with it.  It might also be enough to support your spouse and kids (many CS PhD students have their families with them). So don't worry too much about the stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at GMU, every year we discuss improving the benefits, and especially stipend, for our graduate students.

@@

\begin{commentbox}[Vu:]
  TA and RA at GMU have similar benefits in tuition waiving and insurance.  The college and department will set the rate for a 9-month graduate assistant stipend.  TA, which is paid by the department, will likely be that amount but RA might be higher depending on the stage of the student (1st year vs ABD\footnote{All but dissertation: close to graduate.}) and the prof.
  \tcblower
  Having health insurance is required at many US universities.  Do not assume that you're young and healthy and ignore insurance.  At GMU, and most good CS PhD programs, your GTA or GRA \emph{will always} come with full insurance. In fact, at GMU your spouse/children will get a significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

\begin{commentbox}[Vu:]
  At GMU CS, admitted PhD students have 4 years of GTA guaranteed, and also receive a stipend for the first summer (\autoref{sec:summer-funding}).
\end{commentbox}

Even if you have other funding and do not need a TA, you still should do TA at least once.  This allows you to see what teaching is like, which is especially helpful for a research career where you often give talks and tell people about your work. GMU sometimes has classes that a more senior student can teach.  In that case, you will be paid as GTA or even sometimes as a lecturer.  This is a good opportunity for students to get teaching experience and also get paid more (as a lecturer).

@@

\begin{commentbox}[Vu:]
  PhD applicants at GMU are automatically eligible for a \emph{Presidential Fellowship}.  It is at least as good as GTA but the most important thing is that as a fellowship it is truly free money (i.e., you do not depend on any prof. or TA).  Adcom members nominate applicants for this fellowship and the whole committee will decide.
\end{commentbox}

@@

If your funding source is TA (\autoref{sec:ta}), you typically do not get paid over the summer as there are few courses.
Some CS departments offer some summer funding, but it is typically not guaranteed and might not be a lot. For example, at GMU, we offer summer funding for 1st-year PhD TA students. The amount over the 3-month summer is similar to their monthly stipend (i.e., their 9-month stipend divided by 3 for the 3 summer months).

For RA (\autoref{sec:ta}), it depends on your prof. and their funding. When writing grant proposals, profs. typically include summer funding for their students (\autoref{sec:ra-cost}). However, funding is never guaranteed, e.g., the prof might not get the grant.
For my students, I have been fortunate to have funding to support them over the summer. Over the 3-summer months, I typically pay them 1/3 of their 9-month stipend. I prioritize summer funding for my students because GMU has very good TA resources so they never have to worry about funding during the AY.

@@

  I am writing to inquire about potential research opportunities as a GRA in your group at GMU. Currently I am an undergraduate student in Computer Science at UNIV and plan to graduate in May 2023.

@@

You \emph{do not} need to do a PhD in CS to work in your area of interest. For example, in addition to a traditional CS department, GMU has IST and Cybersecurity departments, both of which have faculty with PhD in CS and work on CS topics (e.g., AI, Security, Robotics).  So it is possible that you still get to do CS research and publish in CS-focused venues even if you're not in a traditional CS program.  It is common to see faculty with CS PhD in a non-CS department as well as faculty with non-CS PhD in a CS department.

@@

\chapter{Miscs and FAQs}
\epigraph{``I want to share something with you – the three little sentences that will get you through life; number 1: Cover for me, number 2: Oh, good idea, Boss, and number 3: It was like that when I got here.''}{\textsc{The Simpsons}}

@@

@@

Before applying, you should talk to your professors and ask them to give you an honest opinion on where you should apply. You should also apply to a range of schools, including ``safety'' ones.

\paragraph{Not a good fit}  You could have an excellent profile, but if you are interested in a research area that the program does not have, you might not be admitted.
Similarly, if no faculty who is willing to advise you (e.g., they are on sabbatical or personal leave, do not have sufficient funding, or already have too many students already), you will not be admitted.  This is good for you, as you don't want to be in a program where no one can advise you.

For example, the 2024 admission cycle has a huge surge in students interested in AI and NLP (thanks, ChatGPT!). Consequently, AI/NLP faculty might be overwhelmed and cannot consider many candidates, even those with excellent profiles.

\paragraph{Low English exam scores (e.g., IELTS or TOEFL)}  While profs. might not care much about these, there might be a certain minimum set by the college or the university that you need to pass to get admitted, especially for TA funding (e.g., you need to communicate well with students as a TA).  Thus, while profs. are willing to argue for your case, they might be reluctant to go against the requirement of the university on standard exams.

Note this is for international students, as domestic ones (and likely those from these \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{countries}) do not have to take these tests.

@@

@@

  %\item \emph{You did not make a good impression}.  For example, you might have contacted a professor and they did not like your email.  Or you might have met a professor (e.g., at a conference) and they did not like your presentation.  This is rare, but it happens.

%

For each application, the GMU application system compiles a single PDF including a summary, transcripts, test scores, LoRs, a CV, SOP, and writing samples. I usually start with the \emph{summary}, checking for low GPAs or test scores below the university minimum. I then skim the \emph{transcripts} for low grades in relevant courses, noting issues like ``many low grades in main courses or unknown international school with good GPA''.  These are not as important as LORs or SOP but I read them first because the review system has questions about them (e.g., ``is GPA good? is IELTS score sufficient?'')

I pay close attention to \emph{strong LoRs} and skim weaker ones, noting either strong letters from well-known professors/researchers or generic content. For the \emph{CV}, I look for publications, research experiences, and notable achievements, noting things like ``published papers in top venues'' or ``gold medal in an international competition.''
I read \emph{SOP} thoroughly (unless it's poorly written then I just skim). I note whether the SOP is exciting, research-oriented, and tailored to our program. All these notes are entered into the evaluation system.

% Around 15 minutes on average (less than 10 mins for definite rejection; 20+ mins for acceptance). In earlier years I spent more time, but now I can quickly tell if an application is good or bad.

% The evaluation system allows for downloading a single pdf that consists of a summary page (e.g., schools, GPA, standard test scores), transcripts, test results, LoRs, CV, SOP, and writing samples (if any). I typically review the application in this order.  I also make notes on what I've seen, using them to support my evaluation.

% I first skim the summary for school, GPA, standard tests and look for red flags like low GPA or test scores that do not meet the minimum requirements set by the university. Then I scan transcripts for low/failing grades on CS and related courses.  For these I make comments such as ``- unknown international school with good GPA, too many low grades on CS and related courses" or ``+ very good GPA from top school XYZ". I do not look at test details as the summary already gives the main scores.

% Then I look at LoRs.  I read carefully good LoRs and quickly skim over bad or generic ones (very easy to tell, you don't even need much experience to distinguish between strong LoRs from the rest). My comments for LoRs include ``+ very strong LoRs from well-known prof. XYZ talks about research experiences'' or ``- generic LoRs, mostly about course work''.  I also note any unusual or unique details in the LoRs, e.g., ``+ LoR mentioned about student's significant contribution on well-known open-source project X''. Sometimes a LoR came directly from a GMU professor who mentioned they already know the student and intend to offer RA, which is a very strong signal for admission (and my comment would be ``Prof. X making RA offer'').

% I skim over CV, looking for paper publications, research work experiences, and other relevant achievements such as winning prizes in well-known international programming/math competitions.  My notes for CV include things such as ``+ has X papers at top venues'' or ``+ gold medal at X''. Sometimes I also note about unusual or unique experiences, e.g., ``+ has a patent on X''.

% Next is the SOP, which I read carefully (but if it is bad then it would be a quick read). My notes for SOP include things such as ``- SOP mostly on course work, no clear motivation, generic" , or ``+ very enjoyable SOP talking about XYZ, customized for GMU, mentioned profs. X,Y,Z".  If the SOP mentioned publications then I would have ``+ has X papers at top venues" or ``+/- has X papers but at unknown places''.  I will not go online to search for the papers (unless they upload them as sample papers).  I also skim over sampling papers, essentially looking over the abstract to see if reads well, and make notes such as ``+ have a paper under submission to unknown venue, but reads well". If the student mentioned a project, I might also look at it (e.g., if it is on Github) and make notes such as ``+ has project on Github with X stars''.

% At this point I have enough to make up my decision, which I enter into the evaluation system. Typically my decisions are either \emph{reject} or \emph{admit with full funding}.  For the latter, I also mention funding sources (e.g., RA from prof. X or TA). I also include my comments to support the decision. Finally, I also recommend very strong students for the University Presidential fellowship (\autoref{sec:fellowships}).

% The adcom chair will look at all evaluations for each application and make a decision (likely following the decision of the reviewers).  If there are discrepancies such as one voting reject but another full funding acceptance (such cases are rare, but do happen), the chair might ask the reviewers to coordinate to reach a consensus.

%\section{Why are rejection letters sent so late?}\label{sec:late-rejection}

@@

@@

Given the high number of quality applicants and limited number of spots, in addition to having a good application profile, you really want to show something that makes you \textbf{stand out}.  For example, even if you do not have research experience, you can talk about your personal projects, as long as they can help show you can do research. For example, if you have an open-source project on Github that is used by many people, has lots of stars in Github, do talk about it. If you often write technical, research-like blogs with many viewers, talk about them too.

@@

@@

In my personal opinion, highlighting URM or LGBTQ+ identity can \emph{be beneficial}, especially if you can articulate how your diverse experiences contribute to diversity and inclusion in academics.  Many universities have a strong commitment to diversity and inclusion and actively
\emph{recruit students from underepresentative backgrounds} (e.g., some \href{https://cse-climate.engin.umich.edu/reports/climate-dei-reports/cse-climate-and-dei-report-2022-2023/#grad-ethnicity}{stats from UMich CSE} and \href{https://www.gmu.edu/news/2022-09/mason-now-top-10-public-university-diversity-innovation-and-cybersecurity-education-us}{GMU} often promotes itself as one of the most diverse universities in the US). Moreover, many scholarships and fellowships are created specifically for URM and LGBTQ+ students, which you should explore if qualify.

@@

Moreover, promoting diversity aligns with the goal of broadening participation in computing (BPC) that is strongly supported within academia (e.g., see the \href{https://plans.bpcnet.org/GeorgeMasonUniversity_ComputerScience_DepartmentalBPCPlan.pdf}{BPC plan at GMU CS}) and government funding agencies such as NSF (e.g., see \href{https://new.nsf.gov/cise/broadening-participation}{here} and \href{https://www.nsf.gov/pubs/2022/nsf22125/nsf22125.jsp}{here}).
Also, sometimes providing this information can make it easier for faculty to apply for certain scholarships or fellowships to support you, e.g., from a private donor that has offered a fellowship only to students from certain backgrounds.

@@

In short, you \emph{do not need} to formally take CS courses, you just need to show that you have this essential knowledge, e.g., through the mentioned ways. Many universities are well aware that incoming graduate students might not have all the technical background, so they often have a \emph{``bridge''} courses to help students catch up.  For example, GMU has four bridge courses corresponding to the four core areas above that incoming students can take to catch up on their CS knowledge.

@@

\autoref{tab:cost} shows the budget breakdown for a GRA per year.
These numbers are based on my experience at public universities in the US. Private universities may have different numbers.  For simplicity, I will assume the department has a 9-month stipend of \$30K and a 3-month summer of \$10K (a third of the 9-month stipend). I will also use GMU tuition rate of about \$15K/year for full-time study (which is quite cheap compared to private universities, e.g., Univ. of Chicago is a whooping \$70K) and a 58.9\% rate on \emph{indirect cost}, which is a typical rate charged for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the students take two conference trips per year, one domestic and one international (conf. registration, airline tickets, taxi, meals, etc are all included).

>>>>

International students, especially those from less well-known universities and smaller countries, often wonder if they should apply to a CS PhD program in the US. Specifically, they are worried about the \textbf{difficulty of getting admitted} and the \textbf{cost of graduate study} in the US. This section addresses these concerns.

First, applying to a good US university \emph{should not} be any harder than at schools in other countries. In fact, it might even be more flexible since CS PhD in the US \emph{do not} require having an MS or a research topic/proposal/adviser in advance. If you believe you have a chance in other countries, e.g., South Korea, Singapore, Germany, UK, Japan, and Australia, then you will surely have a chance in the US as well. \autoref{sec:non-us-differences} compares CS PhD study in the US to other countries.

@@

\begin{commentbox}[Vu:]
  One of the reasons I created this document is that my colleagues at Mason are interested in recruiting Vietnamese students and are surprised when seeing very few applications in Vietnam (see \autoref{sec:ack}). Each year our CS PhD program receives 500+ PhD applications, most of which are international but only 5--6 are from Vietnam. In general the number of CS PhD applications from Vietnam to US universities is very low and more would be welcome.
\end{commentbox}

@@

The guideline can be useful for all students, but it is primarily intended for \textbf{international students} from \textbf{smaller countries and less well-known universities} (or domestic students from US universities with no PhD programs or limited research experiences). 
Students from top schools with strong research cultures and experiences might already know most of the information given here.
My goal is thus to level the playing field by providing information that is not readily available to less privileged students. 
I also hope to encourage more students with such backgrounds to apply to CS PhD programs in the US.

While this document may not fully meet the needs of students aiming for top-10 schools, which typically have more stringent requirements (e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU), it should be valuable for those targeting other reputable R1 institutions. It is certainly useful for students aiming for Mason!

@@

\section*{Additional Resources}
\begin{itemize}
  \item \href{https://parentheticallyspeaking.org/articles/us-cs-phd-faq/}{Getting a Computer Science PhD in the USA} by Shriram Krishnamurthi
  \item  \href{https://pg.ucsd.edu/PhD-application-tips.htm}{Blog post on PhD Admission} by Philip Guo
  \item   \href{https://talkingtorobots.com/FAQ.html}{PhD Admission post} by Yonatan Bisk (customized for CMU admission)
\end{itemize}

@@

After you submit your PhD application (e.g., around December of year $X$ for the Fall admission of year $X+1$), it will first be screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters? Usually, this screening process is done through a central university system, i.e., not by CS faculty. If you pass this screen, your application will be forwarded to the CS department for further evaluation. If you don't,  the system will tell you what is missing and what you need to do. So pay attention to your email and check your application status regularly.

\begin{commentbox}[Hakan:]
  At Mason, for full consideration, students should make sure to submit \textbf{ALL} required documents by the application deadline, and should never assume that some required documents (such as official TOEFL scores or official diplomas/transcripts) will be waived by the admissions office. If something is listed and not marked as ``optional'', it is mandatory and they should plan for submitting all those.
\end{commentbox}

Then, applications are reviewed by a \emph{PhD admission committee} (\textbf{adcom}) that consists of faculty members in CS\footnote{In some cases the committee can involve affiliated faculty from different disciplines.}. These adcom members have a wide range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depend on the department size. At Mason, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review 25--30 applications. Note that most large schools, including Mason, have separate adcoms for MS programs.

@@

At Mason, we usually decide that a full-time PhD candidate is either (i) admitted with funding (\autoref{sec:funding}) or (ii) rejected. In other words, in most cases, we either
admit you with full funding or reject your application. In some rare cases, we admit
without funding because you have funding on your own, e.g.,
supported by your government or having external fellowships. We justify
our decision with a summary of your application, where we list
strengths, e.g., a well-known school, and weaknesses, e.g., weak
LORs.

@@

\begin{commentbox}[Craig:]
  Mason and many other universities allow you to upload your published papers and other writing samples. In many cases, even if the papers were not published at top places, we can still determine their quality by simply skimming over the paper.
\end{commentbox}

@@

\begin{commentbox}[Vu:]
  Sometimes PhD adcom in the US will share a document such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Foreign-Top-Schools}{this one}, which lists the top schools in several countries. We also ask other faculty and students if we think they know about the place.  For example, when I was a postdoc at UMD, members of their CS PhD adcom asked me to evaluate applicants from Vietnam.  During my time at UNL and now here at Mason, I have looked at Vietnamese applications (whether they are assigned to me or not) and provided input to their reviewers, e.g., X is the top tech school in Vietnam and so it should be \emph{top} instead of \emph{unknown} foreign, which makes a huge difference.
\end{commentbox}
\begin{commentbox}[Deepak:]
  If an applicant is anxious about their school not being known outside their country, they can provide information about their school and department, with independent sources where such information can be verified.
\end{commentbox}

@@

\paragraph{English Tests (IELTS, TOEFL)} Unless your degrees are from specific countries such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-Mason#standard-tests-waiver-eligible-countries}{these}, you will need to
take standardized English tests.

@@

\begin{commentbox}[Vu:]
  The minimum for Mason (being above this might not mean much, but below is a \red{red flag}).
  \begin{itemize}
    \item GPA: $\ge 3.0$ in your undergrad (but we also consider the rank/prestige of your school)
    \item GRE: not required
    % \item but if you want to use it, then we expect a total (V+Q) of $\ge 311$ (with a $\ge 157$ Q) and A $\ge 3.0-3.5$.
    \item English requirement tests (one of the below)
          \begin{itemize}
            \item TOEFL: 88 pts in total AND $\ge 20$ points in each subsection OR
            \item IELTS: $\ge 6.5$ OR
            \item DuoLingo Graduate English: $\ge 120$ OR
            \item Pearson Test of Academic English: $\ge 67$
          \end{itemize}
  \end{itemize}
\end{commentbox}

@@

\begin{commentbox}[Vu:]
  At Mason, faculty are encouraged to interview candidates. For very strong candidates, the interview is actually to recruit them.  In some cases a faculty interviews a candidate that they see potential and want to advocate for their admission. Without the interview, such applications may be more likely to be rejected.
  \tcblower
  In short, getting an interview is a good sign; it means that someone is considering you. If we are not interested in your application, we will not proceed with an interview.
\end{commentbox}

@@

\paragraph{Not getting interviews} While it is generally good to get an interview, not getting one \emph{does not} mean you're out.  Many programs do not have the tradition of interviewing applicants. For example, at Mason, most admitted students with TA (\autoref{sec:ta}) do not go through interviews.

@@

\begin{commentbox}[Vu:]
  Mason has \emph{Virtual} Open House (VOH), e.g., \url{https://cs-Mason.github.io/cs-phd-voh-s23/}. We invite all admitted PhD students to the VOH through Zoom to learn about the CS program, the department, Mason, and the DC area in general. Students also get opportunities to chat with professors and current students.
\end{commentbox}

@@

\paragraph{Low Stipend?} Students often complain their stipend is too low. However, in most cases it is not bad, and you can live comfortably with it.  It might also be enough to support your spouse and kids (many CS PhD students have their families with them). So don't worry too much about the stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at Mason, every year we discuss improving the benefits, and especially stipend, for our graduate students.

@@

\begin{commentbox}[Vu:]
  TA and RA at Mason have similar benefits in tuition waiving and insurance.  The college and department will set the rate for a 9-month graduate assistant stipend.  TA, which is paid by the department, will likely be that amount but RA might be higher depending on the stage of the student (1st year vs ABD\footnote{All but dissertation: close to graduate.}) and the prof.
  \tcblower
  Having health insurance is required at many US universities.  Do not assume that you're young and healthy and ignore insurance.  At Mason, and most good CS PhD programs, your GTA or GRA \emph{will always} come with full insurance. In fact, at Mason your spouse/children will get a significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

\begin{commentbox}[Vu:]
  At Mason CS, admitted PhD students have 4 years of GTA guaranteed, and also receive a stipend for the first summer (\autoref{sec:summer-funding}).
\end{commentbox}

Even if you have other funding and do not need a TA, you still should do TA at least once.  This allows you to see what teaching is like, which is especially helpful for a research career where you often give talks and tell people about your work. Mason sometimes has classes that a more senior student can teach.  In that case, you will be paid as GTA or even sometimes as a lecturer.  This is a good opportunity for students to get teaching experience and also get paid more (as a lecturer).

@@

\begin{commentbox}[Vu:]
  PhD applicants at Mason are automatically eligible for a \emph{Presidential Fellowship}.  It is at least as good as GTA but the most important thing is that as a fellowship it is truly free money (i.e., you do not depend on any prof. or TA).  Adcom members nominate applicants for this fellowship and the whole committee will decide.
\end{commentbox}

@@

If your funding source is TA (\autoref{sec:ta}), you typically do not get paid over the summer as there are few courses.
Some CS departments offer some summer funding, but it is typically not guaranteed and might not be a lot. For example, at Mason, we offer summer funding for 1st-year PhD TA students. The amount over the 3-month summer is similar to their monthly stipend (i.e., their 9-month stipend divided by 3 for the 3 summer months).

For RA (\autoref{sec:ta}), it depends on your prof. and their funding. When writing grant proposals, profs. typically include summer funding for their students (\autoref{sec:ra-cost}). However, funding is never guaranteed, e.g., the prof might not get the grant.
For my students, I have been fortunate to have funding to support them over the summer. Over the 3-summer months, I typically pay them 1/3 of their 9-month stipend. I prioritize summer funding for my students because Mason has very good TA resources so they never have to worry about funding during the AY.

@@

  I am writing to inquire about potential research opportunities as a GRA in your group at Mason. Currently I am an undergraduate student in Computer Science at UNIV and plan to graduate in May 2023.

@@

You \emph{do not} need to do a PhD in CS to work in your area of interest. For example, in addition to a traditional CS department, Mason has IST and Cybersecurity departments, both of which have faculty with PhD in CS and work on CS topics (e.g., AI, Security, Robotics).  So it is possible that you still get to do CS research and publish in CS-focused venues even if you're not in a traditional CS program.  It is common to see faculty with CS PhD in a non-CS department as well as faculty with non-CS PhD in a CS department.

@@

\chapter{Miscs and FAQs}\label{sec:faqs}

\epigraph{``I want to share something with you – the three little sentences that will get you through life; number 1: Cover for me, number 2: Oh, good idea, Boss, and number 3: It was like that when I got here.''}{\textsc{The Simpsons}}

@@

\paragraph{Not a good fit and Bad luck}  You could have an excellent profile, but if you are interested in a research area that the program does not have, you might not be admitted.
Similarly, if no faculty who is willing to advise you (e.g., they are on sabbatical or personal leave, do not have sufficient funding, or already have too many students already), you will not be admitted.  This is good for you, as you don't want to be in a program where no one can advise you.

Related to this is that you just have bad luck and apply at the wrong time.  For example, since 2024 there has been a huge surge in students interested in AI and NLP (thanks, ChatGPT!). Consequently, AI/NLP faculty might be overwhelmed and cannot consider many candidates, even those with excellent profiles.

@@

Before applying, you should talk to your professors and ask them to give you an honest opinion on where you should apply. To increase their chances, many students apply to a range of schools, including ``safety'' ones.

@@

\paragraph{Low English exam scores (e.g., IELTS or TOEFL)}  Profs. might not care much about these, but the college or the university often sets a minimum that you need to pass to be considered, especially for TA funding (e.g., you need to communicate well with students as a TA).  Thus, while profs. are willing to argue for your case, they might be reluctant to go against the requirement of the university on standard exams.

Note that English exams are not required for all students, e.g., domestic students and those from certain countries are waived from taking these exams.  For example, at Mason also waive the English exams for students from these \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-Mason#standard-tests-waiver-eligible-countries}{countries}.

@@

\section{Timeline: Why are rejection letters sent so late?}\label{sec:late-rejection}

@@

\paragraph{Acceptance Letters} Universities typically prioritize sending out acceptance letters first. This allows the admitted students to start planning and preparing for their studies. These letters go to the top candidates who are offered admission directly without being placed on a waitlist.  
Some universities have rolling admissions, meaning they send out acceptance letters as they review applications (e.g., some universities sent out letters as soon as early February). Others have a specific date when they send out the first round of acceptance letters.

\paragraph{Response Deadlines}: Accepted students are usually given a strict deadline to respond to their offers (e.g., whether they will accept the offer), often around \textbf{April 15th} in the US. This is a standard deadline for PhD programs due to the Council of Graduate Schools (CGS) Resolution. After this date, programs can gauge how many slots remain unfilled.

\paragraph{Waitlist}: If the CS program has X slots, then many strong candidates will be waitlisted. If accepted students decline the offer, then offers are sent to students on the waitlist. So if you see people getting accepted, that does not mean you are out yet.

\paragraph{Rejection Letters} Typically, schools start sending out rejection letters to remaining applicants after they have finalized their admissions decisions. Thus, rejection letters are often sent out late (e.g., after April 15th or even much later).

@@

Given the high number of quality applicants and a limited number of spots, in addition to having a good application profile, you really want to show something that makes you \textbf{stand out}.  For example, even if you do not have research experience, you can talk about your personal projects, as long as they can help show you can do research. For example, if you have an open-source project that has lots of stars in Github, then you should mention it. If you often write technical, research-like blogs with many viewers, talk about them too.

@@

For each application, the Mason application system compiles a single PDF including a summary, transcripts, test scores, LoRs, a CV, SOP, and writing samples. I usually start with the \emph{summary}, checking for low GPAs or test scores below the university minimum (\autoref{sec:standard-tests}). I then skim the \emph{transcripts} for low grades in relevant courses, noting issues like ``many low grades in main courses or unknown international school with good GPA''.  These are not as important as LORs or SOP but I read them first because the review system has questions about them (e.g., ``is GPA good? is IELTS sufficient?'')

I read carefully \emph{strong LoRs} and skim weaker ones, noting either strong letters from well-known professors/researchers or generic content. For the \emph{CV}, I look for publications, research experiences, and notable achievements, noting things like ``published papers in top venues'' or ``gold medal in an international competition.''
I read strong \emph{SOP} thoroughly (and skim weak ones). I note whether the SOP is exciting, research-oriented, and tailored to our program. All these notes are entered into the evaluation system.

@@

In my personal opinion, highlighting URM or LGBTQ+ identity can \emph{be beneficial}, especially if you can articulate how your diverse experiences contribute to diversity and inclusion in academics.  Many universities have a strong commitment to diversity and inclusion and actively
\emph{recruit students from underepresentative backgrounds} (e.g., some \href{https://cse-climate.engin.umich.edu/reports/climate-dei-reports/cse-climate-and-dei-report-2022-2023/#grad-ethnicity}{stats from UMich CSE} and \href{https://www.Mason.edu/news/2022-09/mason-now-top-10-public-university-diversity-innovation-and-cybersecurity-education-us}{Mason} often promotes itself as one of the most diverse universities in the US). Moreover, many scholarships and fellowships are created specifically for URM and LGBTQ+ students, which you should explore if qualify.

@@

Moreover, promoting diversity aligns with the goal of broadening participation in computing (BPC) that is strongly supported within academia (e.g., see the \href{https://plans.bpcnet.org/GeorgeMasonUniversity_ComputerScience_DepartmentalBPCPlan.pdf}{BPC plan at Mason CS}) and government funding agencies such as NSF (e.g., see \href{https://new.nsf.gov/cise/broadening-participation}{here} and \href{https://www.nsf.gov/pubs/2022/nsf22125/nsf22125.jsp}{here}).
Also, sometimes providing this information can make it easier for faculty to apply for certain scholarships or fellowships to support you, e.g., from a private donor that has offered a fellowship only to students from certain backgrounds.

@@

In short, you \emph{do not need} to formally take CS courses, you just need to show that you have this essential knowledge, e.g., through the mentioned ways. Many universities are well aware that incoming graduate students might not have all the technical background, so they often have a \emph{``bridge''} courses to help students catch up.  For example, Mason has four bridge courses corresponding to the four core areas above that incoming students can take to catch up on their CS knowledge.

@@

\autoref{tab:cost} shows the budget breakdown for a GRA per year.
These numbers are based on my experience at public universities in the US. Private universities may have different numbers.  For simplicity, I will assume the department has a 9-month stipend of \$30K and a 3-month summer of \$10K (a third of the 9-month stipend). I will also use Mason tuition rate of about \$15K/year for full-time study (which is quite cheap compared to private universities, e.g., Univ. of Chicago is a whooping \$70K) and a 58.9\% rate on \emph{indirect cost}, which is a typical rate charged for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the students take two conference trips per year, one domestic and one international (conf. registration, airline tickets, taxi, meals, etc are all included).
